Hierarchical Clustering
========================================================

## 04W03L01-03

Easy example

1. Creating a data and plotting them.

    ```{r}
    set.seed(1234)
    par(mar = c(0, 0, 0, 0))
    x <- rnorm(12, mean = rep(1:3, each = 4), sd = 0.2)
    y <- rnorm(12, mean = rep(c(1, 2, 1), each = 4), sd = 0.2)
    plot(x, y, col = "blue", pch = 19, cex = 2)
    text(x + 0.05, y + 0.05, labels = as.character(1:12))
    
    ```
2. Creating a **data frame** and **DISTANCE**  
    dist(x, method = "euclidean", diag = FALSE, upper = FALSE, p = 2)  
    the distance measure to be used. This must be one of "euclidean", "maximum", "manhattan", "canberra",
    "binary" or "minkowski". Any unambiguous substring can be given.
    
    ```{r}
    dataFrame <- data.frame(x = x, y = y)
    distxy<-dist(dataFrame)
    distxy
    
    ```
3. Plotting dendrogram
    
    ```{r}
    hClustering <- hclust(distxy)
    plot(hClustering)
    
    ```
    
4. Prettier one  
    function:
        
    ```{r}
    myplclust <- function(hclust, lab = hclust$labels, lab.col = rep(1, length(hclust$labels)),
    hang = 0.1, ...) {
    ## modifiction of plclust for plotting hclust objects *in colour*! Copyright
    ## Eva KF Chan 2009 Arguments: hclust: hclust object lab: a character vector
    ## of labels of the leaves of the tree lab.col: colour for the labels;
    ## NA=default device foreground colour hang: as in hclust & plclust Side
    ## effect: A display of hierarchical cluster with coloured leaf labels.
    y <- rep(hclust$height, 2)
    x <- as.numeric(hclust$merge)
    y <- y[which(x < 0)]
    x <- x[which(x < 0)]
    x <- abs(x)
    y <- y[order(x)]
    x <- x[order(x)]
    plot(hclust, labels = FALSE, hang = hang, ...)
    text(x = x, y = y[hclust$order] - (max(hclust$height) * hang), labels = lab[hclust$order],
    col = lab.col[hclust$order], srt = 90, adj = c(1, 0.5), xpd = NA, ...)
    }
    ```
    and its use
        
    ```{r}
    myplclust(hClustering, lab = rep(1:3, each = 4), lab.col = rep(1:3, each = 4))
    ```
    
5. heatmap
    ```{r}
    set.seed(143)
    dataMatrix <- as.matrix(dataFrame)[sample(1:12), ]
    heatmap(dataMatrix)
    ```
## Better dendrograms

[From](http://rstudio-pubs-static.s3.amazonaws.com/1876_df0bf890dd54461f98719b461d987c3d.html) 


```{r}
# prepare hierarchical cluster
hc = hclust(dist(mtcars))
# very simple dendrogram
plot(hc)
# labels at the same level
plot(hc, hang = -1)
```
    
    
```{r}
# tweeking some parameters
op = par(bg = "#DDE3CA")
plot(hc, col = "#487AA1", col.main = "#45ADA8", col.lab = "#7C8071", 
    col.axis = "#F38630", lwd = 3, lty = 3, sub = "", hang = -1, axes = FALSE)
# add axis
axis(side = 2, at = seq(0, 400, 100), col = "#F38630", labels = FALSE, 
    lwd = 2)
# add text in margin
mtext(seq(0, 400, 100), side = 2, at = seq(0, 400, 100), line = 1, 
    col = "#A38630", las = 2)
```

    
```{r}
par(op)

```

Alternatives
```{r}
# using dendrogram objects
hcd = as.dendrogram(hc)
# alternative way to get a dendrogram
plot(hcd)

```

    
```{r}
# using dendrogram objects
plot(hcd, type = "triangle")
```

**ZOOMING**


```{r}
# plot dendrogram with some cuts
op = par(mfrow = c(2, 1))
plot(cut(hcd, h = 75)$upper, main = "Upper tree of cut at h=75")
plot(cut(hcd, h = 75)$lower[[2]], main = "Second branch of lower tree with cut at h=75")
```

**CUSTOMIZE** - function dendrapply

```{r}
# vector of colors labelColors = c('red', 'blue', 'darkgreen', 'darkgrey',
# 'purple')
labelColors = c("#CDB380", "#036564", "#EB6841", "#EDC951")
# cut dendrogram in 4 clusters
clusMember = cutree(hc, 4)
# function to get color labels
colLab <- function(n) {
    if (is.leaf(n)) {
        a <- attributes(n)
        labCol <- labelColors[clusMember[which(names(clusMember) == a$label)]]
        attr(n, "nodePar") <- c(a$nodePar, lab.col = labCol)
    }
    n
}
# using dendrapply
clusDendro = dendrapply(hcd, colLab)
# make plot
plot(clusDendro, main = "Cool Dendrogram", type = "triangle")

```

Phylogenetic

```{r}
# load package ape; remember to install it: install.packages('ape')
library(ape)
# plot basic tree
plot(as.phylo(hc), cex = 0.9, label.offset = 1)
```

another graphs

```{r}
# cladogram
plot(as.phylo(hc), type = "cladogram", cex = 0.9, label.offset = 1)
# unrooted
plot(as.phylo(hc), type = "unrooted")
# fan
plot(as.phylo(hc), type = "fan")
# radial
plot(as.phylo(hc), type = "radial")
# add colors randomly
plot(as.phylo(hc), type = "fan", tip.color = hsv(runif(15, 0.65, 
    0.95), 1, 1, 0.7), edge.color = hsv(runif(10, 0.65, 0.75), 1, 1, 0.7), edge.width = runif(20, 
    0.5, 3), use.edge.length = TRUE, col = "gray80")
# vector of colors
mypal = c("#556270", "#4ECDC4", "#1B676B", "#FF6B6B", "#C44D58")
# cutting dendrogram in 5 clusters
clus5 = cutree(hc, 5)
# plot
op = par(bg = "#E8DDCB")
# Size reflects miles per gallon
plot(as.phylo(hc), type = "fan", tip.color = mypal[clus5], label.offset = 1, 
    cex = log(mtcars$mpg, 10), col = "red")

```

color in leaves

```{r}
# install.packages('sparcl')
library(sparcl)
# colors the leaves of a dendrogram
y = cutree(hc, 3)
ColorDendrogram(hc, y = y, labels = names(y), main = "My Simulated Data", 
    branchlength = 80)

```

ggdendro

```{r}
# install.packages('ggdendro')
library(ggdendro)
# basic option
ggdendrogram(hc)
# another option
ggdendrogram(hc, rotate = TRUE, size = 4, theme_dendro = FALSE, color = "tomato")
# Triangular lines
ddata <- dendro_data(as.dendrogram(hc), type = "triangle")
ggplot(segment(ddata)) + geom_segment(aes(x = x, y = y, xend = xend, 
    yend = yend)) + ylim(-10, 150) + geom_text(data = label(ddata), aes(x = x, 
    y = y, label = label), angle = 90, lineheight = 0)
```

colored

```{r}
# load code of A2R function
source("http://addictedtor.free.fr/packages/A2R/lastVersion/R/code.R")
# colored dendrogram
op = par(bg = "#EFEFEF")
A2Rplot(hc, k = 3, boxes = FALSE, col.up = "gray50", col.down = c("#FF6B6B", 
    "#4ECDC4", "#556270"))
par(op)
# another colored dendrogram
op = par(bg = "gray15")
cols = hsv(c(0.2, 0.57, 0.95), 1, 1, 0.8)
A2Rplot(hc, k = 3, boxes = FALSE, col.up = "gray50", col.down = cols)
```


